---
title: "Non-teaching rate"
author: "Erin Becker"
date: "May 18, 2016"
output: 
  html_document:
    toc: yes
---

```{r global-options, echo = FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      eval = TRUE)
```

## Instructor and trainee involvement in workshops

This analysis is meant to answer questions about the extent of the non-teaching instructor and the 
non-certified trainee issues.

**Analysis and Data Files**

The RMarkdown that generated this document: [Completion_rates.Rmd]()  
The data: In [https://github.com/datacarpentry/metrics/tree/master/instructor_data](https://github.com/datacarpentry/metrics/tree/master/instructor_data) the files [instructor_data_5_17_16_no_ids.csv](https://raw.githubusercontent.com/datacarpentry/metrics/master/instructor_data/instructor_data_5_17_16_no_ids.csv) and [trainee_data_5_18_16_no_ids.csv]("https://raw.githubusercontent.com/datacarpentry/metrics/master/instructor_data/trainee_data_5_18_16_no_ids.csv")

**Questions**  

Questions to be answered include: 

1) What percent of fully certified instructors haven't taught?  
    - Overall  
    - Only those > 1 year past training  
2) What percent of one-time instructors have only taught at their home institution?
3) What percent of trainees haven't finished checkout within 90 days of training?  
    - Including online trainees
    - Excluding online trainees

Data was collected from AMY database (instructor completion) and Google sheet (trainee completion) by `Greg Wilson` on `5/17/16`.

*Note: This data can not answer Question #2.*

Definitions:  
*Trainee* - person who has gone through instructor training but may not have completed checkout.  
*Instructor* - person who has gone through both instructor training and checkout.

## Trainee completion rate

```{r get-trainee-data}
trainees = read.csv("https://raw.githubusercontent.com/datacarpentry/metrics/master/instructor_data/trainee_data_5_18_16_no_ids.csv", stringsAsFactors = FALSE)
trainees$End.Date = as.Date(trainees$End.Date, format = "%m/%d/%Y")
date_pulled = as.Date('2016-5-17')
trainees$Days.Since = date_pulled - trainees$End.Date # differs slightly from Greg's calculated "day's since" - mine counts from completion date
trainees$percent = round(trainees$Completed/trainees$Participants*100,2)
```

```{r over-90-days}
# looking only at trainings from at least 90 days ago
trainees_old = trainees[which(trainees$Days.Since >= 90),]
perc_co_overall = round(sum(trainees_old$Completed)/sum(trainees_old$Participants)*100,2)

# looking only at online trainings
trainees_online = trainees_old[grep("online",trainees_old$Class),]
perc_co_online = round(sum(trainees_online$Completed)/sum(trainees_online$Participants)*100,2)

# looking only at in-person trainings
trainees_in_person = trainees_old[grep("online",trainees_old$Class, invert = TRUE),]
perc_co_in_person = round(sum(trainees_in_person$Completed)/sum(trainees_in_person$Participants)*100,2)
```

Looking only at training sessions from at least 90 days ago:  
- overall completion **`r perc_co_overall`%**  
- in person completion **`r perc_co_in_person`%**  
- online completion **`r perc_co_online`%**. (Two events, 63% and 31%).  

Most (10/14) in person events have >65% completion rate. Some (Arlington, OK, Melbourne, Florida) much lower.

**Takeaway:** Overall, online sessions didn't have a noticably lower completion rate than in-person sessions, but this appears to be due to a few in-person sessions have very low completion rates.  

Some other events do not look on track to meeting normal completion rates (e.g. UCDavis - 84 days, 39%; UW - 68 days, 14%). More follow-up with these participants likely needed.

*Wonder whether these abnormal rates are due to issues with local community, issues with how training session went, or some other factor.*

Summary of completion rates per event:
```{r summary-trainee}
summary(trainees_old$percent)
```
Note that mean completion rate for individual events is **not** the same as mean overall completion rate, as the number of participants per event varies.

## Instructor teaching rate

```{r get-instructor-data}
instructors = read.csv("https://raw.githubusercontent.com/datacarpentry/metrics/master/instructor_data/instructor_data_5_17_16_no_ids.csv", stringsAsFactors=FALSE)
instructors$Date = as.Date(instructors$Date)
instructors$Certified = as.Date(instructors$Certified)
```

```{r percent-not-teaching}
date_pulled = as.Date('2016-5-17') # should be date data pulled
year_ago = date_pulled - 366 # leap year

# instructors trained over a year ago
old_inst = instructors[which(instructors$Certified < year_ago),]
names_old_inst = unique(old_inst$Person)

# non-teaching instructors
all_non_teach = instructors[which(is.na(instructors$Date)),]
names_all_non_teach = unique(all_non_teach$Person)

# non-teaching old instructors
non_teach_old = old_inst[which(is.na(old_inst$Date)),]
names_non_teach_old = unique(non_teach_old$Person)

# percent of instructors trained over 1 year ago that haven't taught
perc_no_teach = round(length(names_non_teach_old)/length(names_old_inst)*100,2)
```

What percent of instructors trained over a year ago haven't yet taught?  
Trained over one year ago: **`r length(names_old_inst)`**  
Of which, haven't taught:  **`r length(names_non_teach_old)`**  
This is **`r perc_no_teach`%**.

What percent of total instructors haven't taught?  
Total trained: **`r length(unique(instructors$Person))`**  
Haven't taught: **`r length(names_all_non_teach)`**  
This is **`r round(length(names_all_non_teach)/length(unique(instructors$Person))*100,2)`%**.

```{r within-last-year}
# instructors trained within last year
new_inst = instructors[which(instructors$Certified > year_ago),]
names_new_inst = unique(new_inst$Person)

# non-teaching new instructors
non_teach_new = new_inst[which(is.na(new_inst$Date)),]
names_non_teach_new = unique(non_teach_new$Person)

# percent of instructors trained in last year who have taught
perc_new_non_teach = round(length(names_non_teach_new)/length(names_new_inst)*100,2)
perc_new_teach = 100-perc_new_non_teach
```

What percent of instructors trained within the past year haven't yet taught?  
Trained within last year: **`r length(names_new_inst)`**  
Of which, haven't taught: **`r length(names_non_teach_new)`**  
This is **`r perc_new_non_teach`%**. (But many of these may have been trained very recently.)

```{r taught-first-year}

# Test whether dates are in ascending order for each instructor
# remove instructors who haven't taught
inst_teach = instructors[which(is.na(instructors$Date) == FALSE),]
uniq_teach = unique(inst_teach$Person) # only teaching instructors
for(i in uniq_teach) {
  rows = inst_teach[which(inst_teach$Person == i),]
  test = identical(sort(rows$Date), rows$Date)
  if(test == FALSE) print(i)
} 

# grab first line for each person (first teaching)
uniq_inst = unique(instructors$Person) # all instructors
first_teaching = data.frame()
for(i in uniq_inst) {
  rows = instructors[which(instructors$Person == i),]
  first_teaching = rbind(rows[1,], first_teaching)
  first_teaching
}

# calculate time between certification and first teaching and add to df
# note some taught before certification
first_teaching$lag = as.numeric(first_teaching$Date - first_teaching$Certified)
first_teaching$num_days = as.numeric(date_pulled - first_teaching$Certified)

# count number who took longer than 1 year to teach
slow_teach = length(which(first_teaching$lag > 366))
no_teach = length(which(is.na(first_teaching$lag) & first_teaching$num_days > 366))
total_opt = length(which(is.na(first_teaching$lag) == FALSE)) + no_teach
```

What percent of instructors teach within their first year?  
Took longer than one year to teach: **`r slow_teach`**  
Haven't taught (been over a year since training): **`r no_teach`**  
Total percent who didn't teach within first year: **`r round(sum(slow_teach + no_teach)/total_opt*100,2)`%**

What is the distribution of time to first teaching? 
```{r plots-time-to-teach}
plot(density(na.exclude(first_teaching$lag)), main = "Days between training and teaching")
```

Note that many (`r (length(which(first_teaching$lag < 0)))`) instructors taught their first workshop before they were officially certified.

Excluding them:  
```{r plots-time-to-teach-pos}
pos_time = first_teaching[which(first_teaching$lag >0),]
plot(density(pos_time$lag), main = "Days between training and teaching (non-retroactive)")
```

Summary of time to teach first workshop (non-retroactive):  
```{r summary-instructors}
summary(pos_time$lag)
```

Are recently trained instructors on track to meet normal teaching rates?   
Overall, half of instructors teach w/in ``r median(pos_time$lag)`` days.

```{r on-track}
# trainees trained between 95-120 days ago
recent_train = first_teaching[which(first_teaching$num_days < 120 & first_teaching$num_days > 95 ),]
recent_no_teach = length(which(is.na(recent_train$lag)))
```

Trained between 95-120 days ago: **`r nrow(recent_train)`**  
Of which, haven't taught: **`r recent_no_teach`**  
This is **`r round(recent_no_teach/nrow(recent_train)*100,2)`%**.  

Recent batch of trainees appear to be on track.  

## Conclusions
Online training sessions do **not** appear to have lower completion rates than in-person sessions. Completion rates are quite variable between training sessions. This may indicate greater follow-up needed. Overall completion rates ~55%. 

Most instructors (~85%) teach within their first year. About half teach within 100 days. Current group of trainees is on track to meet that target.

